---
title: "singcar_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{singcar_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(singcar)
```

# Overview

The aim of the R package `singcar` is to provide and encourage usage of
appropriate statistical methods for comparing a case against a control sample.
For instance, they may commonly be done in a neuropsychological context, in
which an individual has incurred a specific brain injury and we wish to test
whether this damage has led to an impairment of some cognitive function and
whether two different functions are dissociable. For many functions there is
normed data available which the patient can be compared against directly.
However, when this is not possible a control sample estimating the population,
against which we wish to compare the patient, must be used. Both frequentist and
Bayesian methods have been developed to do this, first and foremost by John
Crawford and Paul Garthwaite (Crawford et al., 2011; Crawford & Garthwaite,
2002, 2007, 2005; Crawford & Howell, 1998). It is these methods that singcar
implements. Power calculators for these tests are also provided. Although the
canonical applications for these tests are in Cognitive Neuropsychology or
Clinical Neuropsychology, they are potentially applicable to any circumstance in
which a measure taken from a single individual is to be compared against data
from a normative sample (i.e. a control group). It should be noted that these
statistical methods could also be applied as a general method of outlier
detection in small samples.
