---
documentclass: jss
classoption: nojss
author:
  - name: Jonathan Ö. Rittmo
    affiliation: University of Edinburgh
    # use this syntax to add text on several lines
    address: |
      | Human Cognitive Neuroscience
      | Psychology
    email: \email{j.rittmo@gmail.com}
  - name: Robert D. McIntosh
    affiliation: University of Edinburgh
    address: |
      | Human Cognitive Neuroscience
      | Psychology
    email: r.d.mcintosh@ed.ac.uk
title:
  formatted: "The \\proglang{R} package \\pkg{singcar}: Comparing Single Cases to Small Samples"
  # If you use tex in the formatted title, also supply version without
  plain:     "The R package singcar: Comparing Single Cases to Small Samples"
  # For running headers, if needed
  short:     "\\pkg{singcar}: Comparing Single Cases to Small Samples"
abstract: >
    Comparison of single cases to populations estimated from a sample has many
    potential applications. Historically, one major application is in the field of
    neuropsychology, where single-case statistics may be used to infer whether an
    individual has suffered a significant cognitive deficit as the consequence of a
    brain lesion. One may wish to estimate whether that individual has abnormally
    low performance on some cognitive ability, or if one cognitive ability is
    abnormally discrepant with respect to another cognitive ability. Several
    statistical methods have been developed to test for abnormality on a single
    variate and abnormality of the difference between two variates when a single
    case is compared to a sample, without losing control of the Type I error rate.
    This paper describes some of the main methods and presents a package in which
    they are implemented.
keywords:
  # at least one keyword must be supplied
  formatted: [single case comparisons, small samples, "\\proglang{R}"]
  plain:     [single case comparisons, small samples, R]
bibliography: ref.bib
preamble: >
  \usepackage{amsmath}
  \usepackage{amsfonts}
  \usepackage{longtable}
  \usepackage{caption}
  \usepackage{pbox}
  \usepackage{booktabs}
  \usepackage{cleveref}
  \renewcommand{\eqref}{\Cref}
  \Crefformat{equation}{#2#1#3}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
vignette: >
  %\VignetteIndexEntry{The R package singcar: Comparing Single Cases to Small Samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')

library(singcar)
```

# Introduction

There are many reasons why researchers and clinicians might want to look at
single cases instead of at the average of some group. In certain fields, such as
neuropsychology, this need arises because the pattern of naturally-occurring
brain damage will be unique in each individual case. From a theoretical
perspective, this means that a single patient might be the only available source
of data for a given phenomenon. From a practical, clinical perspective,
diagnosis and description of the pattern of cognitive impairment is done at the
individual level. Individual brain-damaged patients are thus often compared to
the healthy population to assess changes in cognitive functioning. If we want to
assess the patient score on some variate Y, for which we do not know the
population parameters, these must be estimated from a sample. Thus, the
single-case of interest is compared to a control sample. There are many other
areas where the application of such methods could also be useful: for example
studies of uncommon human expertise, targeted quality checks in industries with
limited production output, or animal studies where rearing a large experimental
group might be infeasible.

As it represents the canonical field for the application of these methods, the
nomenclature of neuropsychology will here be adopted. An abnormally low score on
a single variate will be referred to as a \emph{deficit}, an important concept
for clinical and basic neuropsychology alike. For the latter area another
concept is also considered to be of cardinal importance: the ability to test for
an abnormally large discrepancy between two variates. This is referred to as a
\emph{dissociation}, which is taken to provide evidence for some degree of
functional independence between two cognitive abilities. By charting
dissociations, a cognitive architecture of the mind can be theorized
[@shalliceNeuropsychologyMentalStructure1988].

During the last 20 years methods have been developed to estimate and test for
both deficits and dissociations in the single case, while controlling the Type I
error rate, mainly by John Crawford and Paul Garthwaite
[e.g.,@crawfordComparingIndividualTest1998;
@crawfordComparingSingleCase2011; @crawfordComparisonSingleCase2007;
@crawfordInvestigationSingleCase2002; @crawfordTestingSuspectedImpairments2005].
They are available as standalone computer programs, only taking summary data as
input, at \url{https://homepages.abdn.ac.uk/j.crawford/pages/dept/psychom.htm}.
But the majority of these methods have not yet been implemented in any standard
statistical environment. By doing so in the package \pkg{singcar} for the
\proglang{R} environment [@rcoreteamLanguageEnvironmentStatistical2020], our aim
is to encourage and simplify their usage. Further, limiting Type II errors has
not received as much attention as limiting Type I errors in single-case
methodology [@mcintoshPowerCalculationsSinglecase2020]. By including novel tools
for power calculations, our hope is to increase awareness of the inherently low
power in this field as well as to aid in the planning and design of experiments.

The \proglang{R} package \pkg{singcar} contains seven functions to estimate a case's
abnormality compared to a normal population estimated from a small sample, three of them
with regards to a single variate and four with regards to the discrepancy between two variates.
Both frequentist and Bayesian methods are provided, all developed originally by
Crawford and colleagues [@crawfordComparingIndividualTest1998;
@crawfordComparingSingleCase2011; @crawfordComparisonSingleCase2007;
@crawfordInvestigationSingleCase2002; @crawfordTestingSuspectedImpairments2005].
Of special note for psychological research are the
methods allowing the inclusion of covariates [@crawfordComparingSingleCase2011]
using Bayesian regression techniques (Section \@ref(cov)). These methods make matching the control
sample to the case less cumbersome.

In Section \@ref(section2) the implemented methods are described in detail and in
Section \@ref(section3) the package is described and usage exemplified with a
dataset from a recent neuropsychological single case study.


# Comparing a single case to small samples {#section2}

## Background

In the neuropsychological application of the methods to be presented,
the variates of interest will be scores obtained by participants on tasks
assessing relevant cognitive functions. The variates will
thus often be referred to as task scores. Historically, it was not uncommon for
researchers to compare single cases against a control
population estimated from a sample by evaluating the case score as a
\(Z\) score from the estimated distribution. The \(p\) value associated with this
\(Z\) score would then be treated as an estimate of the case's abnormality. A
similar logic was sometimes applied for estimating the abnormality of the
discrepancy between two tasks, using a method developed by \citet{payneStatisticsInvestigationIndividual1957}:
researchers calculated a \(Z\) score based on the case's difference
between two standardised variates divided by the standard deviation of the
difference.

The \(Z\) score approach is of course problematic because it treats parameter estimations from
a restricted control sample as if they were population parameters. This could
be appropriate with very large samples, but
control samples in neuropsychology are often small (sometimes \textless{} 10), and it is
well known that the sampling distribution of the sample variance is right skewed
for small sample sizes. Hence, underestimation of variance would be more
probable than overestimation, inflating obtained \(Z\) scores and thus
Type I errors if the Z scores were used for hypothesis testing \citep{crawfordComparingIndividualTest1998}.

The following sections describe methods that have been devised to
allow for the evaluation of the abnormality of a case against a restricted
control sample, whilst retaining appropriate control over the Type I error rate.
These methods include frequentist approaches (Section \ref{sec22}) and Bayesian
approaches (Section \ref{sec23}).


## Frequentist approaches {#sec22}

An appropriate method for comparing a single observation to the mean of a
sample was proposed within the biological sciences by \citet{sokalBiometryPrinciplesPractice1981} (p.
227). It was popularized within neuropsychology by \citet{crawfordComparingIndividualTest1998},
where its common application was as a (one-tailed) test of deficit (TD) to
determine whether the score of a single case with brain damage was abnormally
low (assuming that a low score indicates poorer performance) with respect to the
mean score of a control sample. The \(t\) distribution is used to account for
the underestimation of the sample variance. The basic approach is a modified two
samples \(t\) test where the case simply is treated as a sample of size 1. The
degrees of freedom for this distribution is \(n + 1 - 2 = n - 1\).

\begin{equation}
t_{n-1} = \frac{y^* - \overline{y}}{s \sqrt{\frac{n + 1}{n}}}
\label{eq:TD}
\end{equation}

Where \(y^*\) is the case score, \(\overline{y}, \ s\) the sample mean and sample
standard deviation respectively and \(n\) the size of the control sample. This
method does not allow estimation of a notional patient population. However,
since the question posed when using this test is about
the probability of the case being part of the control population, the potential alternative
affinity of the case is of no concern \citep{crawfordComparingIndividualTest1998}. This
test of deficit provides transparent control of the Type I error rate \citep{crawfordComparingSingleCase2009, crawfordInferentialMethodsComparing2004, crawfordSinglecaseResearchNeuropsychology2012}.
Moreover, the \(p\) value obtained constitutes an unbiased point estimate of the
abnormality of the case, as demonstrated by \citet{crawfordMethodsTestingDeficit2006}. The
simple proof for this is given below.

The proportion of controls that would score lower on a random variable \(Y\) than
the case \(y^*\) is
\[
\mathbb{P}[Y< y^*]
\]
Subtracting \(\overline{y}\) from both sides of the inequality
and dividing by \(s \sqrt{\frac{n + 1}{n}}\)
\[
\mathbb{P}[Y< y^*] =\mathbb{P}\left[\frac{Y-\overline{y}}{s \sqrt{\frac{n + 1}{n}}} < \frac{y^* - \overline{y}}{s \sqrt{\frac{n + 1}{n}}}\right]
\]
The quantity to the left of the inequality, i.e., \(\frac{y-\overline{y}}{s \sqrt{\frac{n + 1}{n}}}\)
is \(t\) distributed with \(n-1\) degrees of freedom. Hence,
\[
\mathbb{P}[Y< y^*] =\mathbb{P}\left[t_{n-1} < \frac{y^* - \overline{y}}{s \sqrt{\frac{n + 1}{n}}}\right]
\]
The quantity to the right of the inequality is the test statistic from
Equation \eqref{eq:TD}, hence \(\mathbb{P}[y< y^*]\) is the same as the \(p\) value obtained
from the test of deficit. This fact also makes the construction of confidence
intervals of this abnormality possible.

\citet{crawfordPointIntervalEstimates2010} pointed out that, although the \(Z\) score
is not appropriate for estimating the abnormality of a case when the control
sample is small, it does provide a standardised effect size measure of
abnormality, similar to Cohen's d \citep{cohenStatisticalPowerAnalysis1988}. Insensitivity to sample size is
indeed a requirement for an effect size index and the proposed quantity was:
\begin{equation}
Z_{CC} = \frac{y^* - \overline{y}}{s}
\label{eq:zcc}
\end{equation}
This is simply a \(Z\) score calculated from sample estimates. The
subscript (CC) indicates that it relates to a "case-controls" comparison. The
construction of confidence intervals for the point estimate of abnormality using
the \(Z_{CC}\) index of effect size was described by \citet{crawfordInvestigationSingleCase2002}.

Put \(p\) as the percentage of the population that would fall below a
case score, then the \(1-\frac{\alpha}{2}\) confidence interval for \(p\) is
constructed as follows: Let \(Z_{CC} = \frac{y^*-\overline{y}}{s}\) and \(y^* \neq \overline{y}\)
then \(Z_{CC}\) comes from a non-central \(t\) distribution on \(n-1\)
degrees of freedom. By deploying a search algorithm we find the value \(\delta_U\)
for the non-centrality parameter (NCP) of a non-central \(t\) distribution on \(n-1\)
degrees of freedom such that the \(100\frac{\alpha}{2}\) percentile equates
\(Z_{CC}\sqrt{n}\) and similarly we find the NCP \(\delta_L\) of a non-central
\(t\) distribution such that its \(100(1-\frac{\alpha}{2})\) percentile equates
\(Z_{CC}\sqrt{n}\). The upper and lower boundaries for \(Z_{CC}\) are then given by:
\[
Z_{CC_U} = \frac{\delta_U}{n}, \ \ Z_{CC_L} = \frac{\delta_L}{n}
\]
and the boundaries for \(p\) by:
\[
p_U = \Phi\left(\frac{\delta_U}{n}\right), \ p_L = \Phi\left(\frac{\delta_L}{n}\right)
\]
Where \(\Phi\) is the CDF of the standard normal distribution. Note that the above
equation is appropriate when the case score falls on the left side of the
distribution. If it were to fall on the right side, then the upper and lower
boundaries would be given by:
\[
p_U = 1 - \Phi\left(\frac{\delta_L}{n}\right), \ p_L = 1 - \Phi\left(\frac{\delta_U}{n}\right)
\]

As has been shown, estimating abnormality on a single variate is a simple
matter. For estimating abnormality of discrepancy, when the normally distributed
variates \(Y_1\) and \(Y_2\) can be compared without standardisation, the approach
is equally simple and in fact identical to the test of deficit, Equation \eqref{eq:TD},
except that the statistic is based on difference scores, taking the correlation
between the variates into account (\(\rho_{12}\)).
\begin{equation}
t_{n-1} = \frac{(y^*_1 - \overline{y}_1) - (y^* _2 - \overline{y}_2) }{ \sqrt{(s^2_1 +s^2_2 -2s_1 s_2 \rho_{12})(\frac{n+1}{n})}}
\label{eq:UDT}
\end{equation}
Construction of confidence intervals for abnormality estimates based on this
unstandardised difference test (UDT) is done in an analogous way to that of the
confidence intervals of the TD \citep{crawfordTestingSuspectedImpairments2005}.
However, the test is somewhat limited in its usefulness, because it is only
applicable if the two variates are measured on equivalent scales. Far more
common, at least in neuropsychology, is the need to assess discrepancies between
differently-scaled variates, which require standardisation to be comparable.

By standardising the variates (not taking sample size into account)
we get an effect size measure similar to \(Z_{CC}\), 
Equation \eqref{eq:zcc} \citep{crawfordPointIntervalEstimates2010}:
\begin{equation}
Z_{DCC}  = \frac{z^*_1 - z^*_2}{\sqrt{2-2\rho_{12}}}
\label{eq:PJ}
\end{equation}
Where \(z_1^*\) and \(z_2^*\) are the standardised case scores on some task A
and some task B and the subscript (DCC) indicates "discrepancy-case-controls".
This quantity was proposed by \citet{payneStatisticsInvestigationIndividual1957}
as a significance test for discrepancies, but of course it is not appropriate
for such a purpose if small samples are used: we cannot use the \(Z\)
distribution for estimating the abnormality of the discrepancy for the same
reason we cannot use \(Z_{CC}\) to test for a deficit.
<!-- But also because we cannot use  \(Z\) scores to represent the case's scores -->
<!-- on each variate since the size of these will be inflated as well. -->

When the case scores on variates \(Y_1\) and \(Y_2\) are estimated from small
samples and they need to be standardised, it means that instead of estimating
the discrepancy between two normally distributed variates we need to estimate
the discrepancy between two \(t\) distributed variates.
Because linear combinations of correlated \(t\) distributed random variates are not
themselves \(t\) distributed this problem is non-trivial. The distribution of such
difference scores was examined by \citet{garthwaiteDistributionDifferenceTwo2004}. They used
asymptotic expansion to find functions of the sample correlation that, if used
as a denominator to the difference between the variates, would yield an
asymptotically \(t\) distributed quantity. They found:

\begin{equation}
\psi=\frac{\frac{(y^*_1-\overline{y}_1)}{s_{1}}-\frac{(y^*_2-\overline{y}_2)}{s_{2}}}{
\sqrt{
(\frac{n+1}{n})
\left( (2-2 \rho)+
\frac{2(1-\rho^{2})}{n-1}+
\frac{(5+c^{2})(1-\rho^{2})}{2(n-1)^{2}}+
\frac{\rho(1+c^{2})(1-\rho^{2})}{2(n-1)^{2}}\right)
}}
\end{equation}

Where \(\rho\) is the sample correlation between the tasks and \(c\) the critical
two-tailed \(t\) value with \(n-1\) degrees of freedom. \citet{garthwaiteDistributionDifferenceTwo2004}
demonstrated that \(\mathbb{P}[ \psi > c] \approx \mathbb{P}[t >c]\). To obtain a
precise probability for \(\psi\), one solves for \(\psi = c\), which gives a quantity
not dependent on a pre-specified critical value.
\(\psi = c\) is a quadratic equation in \(c^2\), choosing the positive root of which
yields:

\begin{align}
\begin{split}
c & = \sqrt{\frac{ -b + \sqrt{b^2 - 4ad}}{2a}}, \  \text{where} \\
a & = (1+r)(1-r^2), \\
b & =  (1-r)[4(n-1)^2+4(1+r)(n-1)+(1+r)(5+r)], \\
d & =  - 2\left[\frac{y^*_{1} - \overline{y}_1}{s_1}-\frac{y^*_2 -\overline{y}_2}{s_2}\right]^2\left(\frac{n(n-1)^2}{n+1}\right)
\end{split}
\label{eq:RSDT}
\end{align}

Where \(p = \mathbb{P}[t_{n-1}>c]\) is the estimate of abnormality, and is used for
significance testing. It should be noted that because \(c\) is quadratic and we
choose the positive root, the resultant statistic cannot be negative. 
If it is required that the test statistic expresses the direction of a negative
effect the correct sign must be imposed.

The quantity in Equation \eqref{eq:RSDT} is referred to as the "revised
standardised difference test"
\citep[RSDT:][]{crawfordTestingSuspectedImpairments2005}. The name stems from it
being preceded by a test similar to that of the unstandardised difference test
(Equation \eqref{eq:UDT}) but with standardised normal variates, developed by
\citet{crawfordPayneJonesRevisited1998}.
\citet{crawfordTestingSuspectedImpairments2005} show with Monte Carlo
simulations that the RSDT is superior in controlling Type I errors compared to
their earlier $t$ test and the $Z$ score method of
\citet{payneStatisticsInvestigationIndividual1957}.

Even for very small sample sizes of \(n=5\), RSDT was shown to barely exceed the
specified 5\% error rate. However, if a case lacks a true discrepancy
between the variates but exhibits extreme scores on both of them (in the same direction), the error rate
of the RSDT increases steeply. \citet{crawfordComparisonSingleCase2007} 
showed that the RSDT starts to lose control of the
error for task scores being more extreme than two standard deviations away from
the mean on both variates, without them exhibiting a discrepancy. For task
scores at 8 standard deviations from the mean, the Type I error rate of the RSDT
was inflated to nearly 35\%.

Another major drawback of the RSDT is that it has proved difficult to construct
confidence limits on the point estimate of abnormality due to \(\psi\) only being
approximately \(t\) distributed. To remedy this and to be able to provide an exact
rather than just approximate estimate of abnormality Crawford and colleagues started
looking into Bayesian methodology.

## Bayesian approaches {#sec23}

In the frequentist framework, parameters are treated as fixed attributes of a
population and their estimations are thought to converge to the true value
across a series of trials. In the Bayesian framework, by contrast, parameters
are treated as random variables with associated probability distributions. To
estimate a parameter distribution, we can use any prior knowledge of that
parameter to assign probabilities to possible values of the parameter, forming
what is known as a prior distribution, or simply a prior. If no prior
information is available we may want to use a non-informative prior, the most
simple of which would assign equal probabilities to all possible parameter
values. The prior is updated when new information is obtained to form what is
called a posterior distribution. The posterior probability of a hypothesis
(i.e., a specified value of the parameter) is calculated by using Bayes theorem.
If we disregard the marginal probability of the data in Bayes theorem it can be
rewritten as:
\begin{equation*}
posterior \ \propto \ likelihood \times prior
\end{equation*}

What this is saying is that the posterior density of a hypothesis is
proportional (\(\propto\)) to the likelihood of the data under that hypothesis times the
prior probability of the hypothesis.

In many cases it is not feasible to calculate the posterior analytically and
instead Monte Carlo methods are often used. These methods are
mathematical algorithms that solve numerical problems by repeated sampling from
distributions. The algorithms used differ depending on the problem at hand, but
in general they are all building on rules of drawing random numbers based on the
likelihood and the prior, and observing the distribution formed after a large
number of iterations. The peak of such a distribution is often taken as an
estimation of the parameter of interest and when using non-informative priors
this often also corresponds to the maximum likelihood estimate. Hence, using a
non-informative prior often yields estimations with frequentist properties, a
feature that is useful if hypothesis testing is desired.

This was a requirement when \citet{crawfordComparisonSingleCase2007} and
\citet{crawfordComparingSingleCase2011} developed Bayesian versions of the tests
described in Section \ref{sec22}, now implemented in \pkg{singcar}. The
following sections describe the procedural details of these methods which are
all based on Monte Carlo simulations. The steps presented closely follow those
outlined in the original papers with just a few slight changes of notation.


### The Bayesian test of deficit {#BTD}

The Bayesian test of deficit (BTD) allows us to obtain an estimate of \(p\) (and accompanying credible
intervals), which is the proportion of controls that would obtain a value more
extreme than the case on the variate of interest.

Assume a sample of \(n\) controls on which we measure
some value \(y\) that is normally distributed with unknown mean \(\mu\) and unknown
variance \(\sigma^2\). Let \(\overline{y}\) and \(s^2\) denote the sample mean and
sample variance respectively and \(y^*\) the case score.
Because \(\mu\) and \(\sigma^2\) are unknown, the prior distribution of \(\mu\) is
conditioned on the prior distribution of \(\sigma^2\).
A non-informative prior \(\mu | \sigma^2 \sim \mathcal{N}(0, \ \infty)\) is
assumed. For the posterior we have that
the marginal distribution of \(\frac{(n-1)s^2}{\sigma^2} \sim \chi^2_{n-1}\) and
the posterior distribution for
\(\mu|\sigma^2 \sim \mathcal{N}(\overline{y}, \sigma^2/n)\), see e.g., \citet{gelmanBayesianDataAnalysis2013} (p. 45 and 65).
To estimate \(p\), the following steps are iterated:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Let \(\psi\) be a random draw from a \(\chi^2\)-distribution on \(n-1 \ df\).
  Then let \(\hat{\sigma}^2_{(i)} = \frac{(n-1)s^2}{\psi}\) be the estimation of
  \(\sigma^2\) for this iteration, hence the subscript \((i)\).
\item
  Let \(z\) be a random draw from a standard normal distribution.
  Then let \(\hat{\mu}_{(i)}=\overline{y}+z\sqrt{(\hat{\sigma}_{(i)}^2/n)}\) be
  the estimate of \(\mu\) for this iteration.
\item
  With estimates of \(\mu\) and \(\sigma\), \(p\) is calculated conditional
  on these estimates being the "correct" \(\mu\) and \(\sigma\). Let
  \(z^*_{(i)}= \frac{y^* - \hat{\mu}_{(i)}}{\sqrt{\hat{\sigma}_{(i)}^2}}\) be the standardised
  case score then \(\hat{p}_{(i)} =\Phi\left(z^*_{(i)}\right)\) or \(\hat{p}_{(i)} = 1-\Phi\left(z^*_{(i)}\right)\)
  depending on alternative hypothesis, is the estimate of \(p\) for this
  iteration. That is the probability of drawing a
  value more extreme than \(z^*_{(i)}\) from a standard normal distribution.
\end{enumerate}

Repeating these steps a large number of times will yield a distribution
of \(\hat{p}\), the mean of which is taken as the point estimate of \(p\) and used
for hypothesis testing. Multiplied by 100 it gives a point estimate of the percentage
of the control population expected to exhibit a more extreme score than the case. If repeated
e.g., 1000 times, the 25th smallest and 25th largest \(\hat{p}_{(i)}\) would give the lower and
upper boundaries of the 95\% credible interval for \(p\). Similarly,
the 25th smallest and 25th largest values of \(z^*_{(i)}\) would give the lower and
upper boundaries of the 95\% credible interval for \(Z_{CC}\) the point
estimate of which is that of Equation \eqref{eq:zcc}. \citet{crawfordComparisonSingleCase2007}
show that this method yields converging results to that of the frequentist test of deficit, Equation \eqref{eq:TD}.

### The Bayesian standardised difference test {#BSDT}

The Bayesian standardised difference test (BSDT) follow a similar procedure to
that of BTD. However, we now assume a sample of \(n\) controls from which we
obtain values on the variates \(Y_1\) and \(Y_2\) following a bivariate normal
distribution and representing task A and task B. Let \(\overline{y}_1\) and
\(\overline{y}_2\) denote the sample means and
\begin{equation*}
\pmb{A}=\begin{bmatrix}
s^2_{1} & s_{12} \\
s_{12} & s^2_{2} \end{bmatrix}
\end{equation*}
the sample variance-covariance matrix, where \(s_{12}\) is the sample covariance,
then \(\pmb{S} =\pmb{A}(n-1)\) is the sums of squares and cross products (SSCP) matrix. It
is assumed that the observations come from a bivariate normal distribution with
unknown mean \(\pmb{\mu}\) and unknown variance \(\Sigma\):

\begin{equation*}
\pmb{\mu} = \begin{pmatrix}
\mu_1 \\
\mu_2 \end{pmatrix} \ \text{and} \ \Sigma=\begin{bmatrix}
\sigma^2_{1} & \sigma_{12} \\
\sigma_{12} & \sigma^2_{2} \end{bmatrix}
\end{equation*}

Let the case scores be denoted \(y_1^*\) and \(y_2^*\). Just as for the frequentist
dissociation tests we want to estimate the proportion \(p\) of the control
population that would exhibit a greater difference \(Y_1-Y_2\) than the case's
\(y_1^*-y_2^*\).

The multivariate generalization of the \(\chi^2\)-distribution is the Wishart
distribution. That is, a Wishart distribution of 1 dimension is a
\(\chi^2\)-distribution on \(n\) degrees of freedom. One can view the Wishart
distribution parameterised with \(n\) degrees of freedom and some
variance-covariance matrix as being a distribution of SSCP-matrices related to
that variance-covariance matrix. Similarly, one can view the inverse Wishart
distribution parameterised with the same degrees of freedom and a SSCP-matrix as
being a distribution of variance-covariance matrices related to that
SSCP-matrix.

The non-informative prior for \(f(\pmb\mu, \Sigma^{-1})\) chosen in
\citet{crawfordComparisonSingleCase2007} was \(f(\pmb\mu, \Sigma^{-1}) \propto
|\Sigma|\) in favour of the perhaps more commonly used \(f(\pmb\mu, \Sigma^{-1})
\propto |\Sigma|^{(k+1)/2}\) ($k =$ number of variates) because, for unstandardised differences, it was shown
to yield identical interval estimates to the frequentist UDT. The posterior
marginal distribution of \(\Sigma^{-1}\) for this choice of prior is a Wishart
distribution with \(n\) degrees of freedom and scale matrix \(\pmb{S}^{-1}\).
The conditional distribution of \(\pmb\mu|\Sigma\) is then a multivariate normal
distribution with mean \([\overline{y}_1 \ \overline{y}_2]\) and variance
\(\Sigma/n\), see e.g., \citet{gelmanBayesianDataAnalysis2013} (p. 72-73).
\citet{crawfordComparisonSingleCase2007} refer to this as the "standard theory"
prior.

Using the standard theory prior produces good frequentist properties for
\(\sigma_1\) and \(\sigma_2\), but \citet{bergerObjectivePriorsBivariate2008}
have shown that the convergence to frequentist estimates is less good for
\(\rho\) and differences in means ($\mu_1 - \mu_2$). Instead, they recommend
that \(f(\pmb\mu, \Sigma^{-1}) \propto \frac{1}{\sigma_1\sigma_2(1-\rho^2)}\)
should be used as a general purpose prior for bivariate normal data.
To construct posteriors with this prior rejection sampling is used. Random
observations are drawn from an inverse Wishart on \(n-1\)
degrees of freedom and only a subset of the draws are accepted.
\citet{crawfordComparingSingleCase2011} noticed however that this prior gave rise to too narrow
credible intervals and, with simulations, showed that if the sample size was
treated as \(n-1\) for estimations of \(\Sigma\), so that the intervals were
somewhat conservative, the frequentist properties of their estimations were
improved. \citet{crawfordComparingSingleCase2011} recommend this and refer to it as the
"calibrated" prior. The procedure for sampling from the posterior using this
prior is given below:


1. Since the unbiased estimate of \(\Sigma\) is \(\pmb{S}/(n-1)\),
   we put \(\pmb{S}^*= (n-2)\pmb{S}/(n-1)\) to retain this property
   when reducing the sample size.

2. Generate an observation from an inverse Wishart distribution
   on \(n-2\) degrees of freedom and with scale matrix \(\pmb{S}^*\).
   Denote this:
   \[
   \hat{\Sigma} = \begin{bmatrix}
   \hat{\sigma}^2_{1} & \hat{\sigma}_{12} \\
   \hat{\sigma}_{12} & \hat{\sigma}^2_{2} \end{bmatrix}
   \]
   and put
   \[
   \hat{\rho}= \frac{\hat{\sigma}_{12}}{\sqrt{\hat{\sigma}^2_{1}\hat{\sigma}^2_{2}}}
   \]

3. Generate a value \(u\) from a uniform distribution such that \(u \sim U(0, 1)\).
   If \(u^2 \leq 1-\hat{\rho}^2\) we accept \(\hat\Sigma\) as the estimation of
   \(\Sigma\) for this iteration and denote it
   \begin{equation*}
   \hat{\Sigma}_{(i)}=\begin{bmatrix}
   \hat{\sigma}^2_{1(i)} & \hat{\sigma}_{12(i)} \\
   \hat{\sigma}_{12(i)} & \hat{\sigma}^2_{2(i)} \end{bmatrix}
   \end{equation*}
   otherwise
   we iterate the procedure until we have an accepted \(\hat{\Sigma}\).


Bayesians might argue that good frequentist properties are not necessary when
conducting Bayesian analyses. If so, one can use the "standard theory" prior
by generating a random observation from an inverse Wishart distribution on \(n\)
degrees of freedom and scale matrix \(\pmb{S}\) and denote it
\(\hat{\Sigma}_{(i)}\). With an estimate of \(\Sigma\) (regardless of the prior
chosen), follow the steps below to obtain an estimate of \(p\), that is the percentage
of the control population expected to show a more extreme score than the case.


1.  Let \(z_{r1}\) and \(z_{r2}\) be two random draws from a standard normal distribution.
    Perform Cholesky decomposition on \(\hat{\Sigma}_{(i)}\), that is finding the lower triangular matrix \(\pmb{T}\)
    such that \(\pmb{T}\pmb{T'}=\hat{\Sigma}_{(i)}\). Then
    \begin{equation*}
    \pmb{\hat{\mu}}_{(i)} = \begin{pmatrix}
    \hat{\mu}_{1(i)} \\
    \hat{\mu}_{2(i)} \end{pmatrix} = \begin{pmatrix}
    \overline{y}_1 \\
    \overline{y}_2 \end{pmatrix}+ \pmb{T} \begin{pmatrix}
    z_{r1} \\
    z_{r2} \end{pmatrix} / \sqrt{n}
    \end{equation*}
    is the estimation of \(\pmb{\mu}\) for this iteration.

2.  With estimations of \(\pmb{\mu}\) and \(\Sigma\) we can calculate \(p\), given that
    they are the the correct values of \(\pmb{\mu}\) and \(\Sigma\). If an unstandardised
    test is required put:
    \begin{equation*}
    z_{(i)}^* = \frac{(y_1^* - \hat{\mu}_{1(i)}) - (y^*_2 - \hat{\mu}_{2(i)})}
    {\sqrt{\hat{\sigma}^2_{1(i)}+\hat{\sigma}^2_{2(i)}-2\hat{\sigma}_{12(i)}}}
    \end{equation*}
    If a standardised test is required\footnote{In the typical application a standardised test will most often be
      required because different cognitive functions frequently are measured on different operationalised scales.}, put:
    \begin{equation*}
    z_{1(i)} = \frac{y_1^* - \hat{\mu}_{1(i)}}{\sqrt{\hat{\sigma}^2_{1(i)}}}, \ z_{2(i)} = \frac{y_2^* -
    \hat{\mu}_{2(i)}}{\sqrt{\hat{\sigma}^2_{2(i)}}}, \ \hat{\rho}_{(i)} = \frac{\hat{\sigma}_{12(i)}}{\sqrt{\hat{\sigma}_{1(i)}\hat{\sigma}_{2(i)}}}
    \end{equation*}
    and
    \begin{equation*}
    z^*_{(i)} = \frac{z_{1(i)} - z_{2(i)}}{\sqrt{2-2\hat{\rho}_{(i)}}}
    \end{equation*}

3.  Let \(\hat{p}_{(i)}\) be the tail area of a standard normal distribution
    less or greater than \(z^*_{(i)}\) (depending on alternative hypothesis).
    \(\hat{p}_{(i)}\) is then the estimate of \(p\) for this iteration.


Repeating these
steps a large number of times will yield a distribution of \(\hat{p}\), the
mean of which is taken as the point estimate of \(p\) and used for hypothesis testing.
Multiplied by 100 it gives the point estimate of the percentage of controls expected to exhibit
a more extreme task difference than the case. If repeated e.g., 1000 times, the
25th smallest and 25th largest \(\hat{p}_i\) is the lower and upper boundaries of the 95\%
credible interval for \(p\). Similarly,
the 25th smallest and 25th largest values of \(z^*_{(i)}\) gives the lower and
upper boundaries of the 95\% credible interval for \(Z_{DCC}\), the point
estimate of which is that of Equation \eqref{eq:PJ}.

### Bayesian tests allowing for covariates {#cov}

\citet{crawfordComparingSingleCase2011} extended the previously described Bayesian tests by
using Bayesian regression techniques that allowed the case's abnormality to be assessed
in the presence of covariates. These tests thus allow you to compare
the case's score on the task of interest conditioned on the results of the
controls having the same score as the case on the covariate(s). If a case has 15
years of education, his/her score on the task would be compared to the controls
with equal length of education.

In addition to improving the precision of these tests, by accounting for
spurious noise, this also facilitates the collection of larger control samples,
because it reduces the need to very closely match control samples to a single
case. Moreover, it means that the same control sample may be used for the
evaluation of multiple single-cases, since the matching on relevant covariates
of interest can be achieved statistically. One degree of freedom is lost for
each covariate included, so as a rule of thumb they should be included only if
they have a sample correlation with the variate(s) of interest stronger than
\(0.3\) \citep{crawfordComparingSingleCase2011}. Of course, the control sample should also ideally
bracket the case on the covariates, in order to avoid extrapolating "out
of sample". As for the earlier tests, the assumption of the variates of interest
are still that they follow a normal or bivariate normal distribution. However,
no assumptions are made about the distribution of the covariates.

Suppose we have data from a sample of size \(n\), with scores on \(m\) covariates
and \(k = 1, 2\) variates of interest, depending on whether we are testing for a
deficit or a discrepancy. We denote the covariates
\(\boldsymbol{X} = (X_1, \dots, X_m)\).

From these values we wish to estimate
\[
\boldsymbol{B} = \begin{bmatrix}
\boldsymbol{\beta}_1 & \cdots & \boldsymbol{\beta}_k
\end{bmatrix}
\]
Where \(\boldsymbol{\beta}_i\) is a vector of length \(m+1\) containing regression
coefficients for each covariate on the \(i\)th variate of interest, the first element
in \(\boldsymbol{\beta}_i\) being the intercept. We also
wish to estimate
\[
\Sigma \ | \ \boldsymbol{X} =
\begin{bmatrix}
\sigma^2_1 \ | \ \boldsymbol{X} & \rho\sigma_1\sigma_2 \ | \ \boldsymbol{X} \\
\rho\sigma_1\sigma_2 \ | \ \boldsymbol{X} & \sigma^2_2 \ | \ \boldsymbol{X}
\end{bmatrix}
\]
That is, the covariance matrix of the variates of interest conditioned upon the
covariates, for task \(Y_1\) and task \(Y_2\) i.e., when \(k=2\). If we wish to test for
a deficit, that is \(k = 1\) then \(\Sigma \ | \ \boldsymbol{X}\) is a \(1\times1\)
matrix containing the conditional variance of the variate of interest.
We assume \(\Sigma\) not to vary with the covariates.

The procedure will be outlined in the general case, that is for \(k = 1\)
or \(k = 2\). The main difference between testing for abnormality on a single
variate and testing for abnormal discrepancy between two variates is the recommended
specification of the prior.

Let \(\boldsymbol{X}\) be the \(n \times m+1\) design matrix on which we regress
\(\boldsymbol{Y}\), the \(n \times k\) response matrix.
\[
\boldsymbol{X} =
\begin{bmatrix}
1 & x_{11} & \cdots & x_{1m} \\
\vdots & \vdots & \ddots & \vdots \\
1 & x_{n1} & \cdots & x_{nm}
\end{bmatrix},\ \
\boldsymbol{Y} =
\begin{bmatrix}
 y_{11} & \cdots & y_{1k} \\
 \vdots & \ddots & \vdots \\
 y_{n1} & \cdots & y_{nk}
\end{bmatrix}
\]
The data estimates of \(\boldsymbol{B}\) and \(\Sigma\) are then
\[
\boldsymbol{B}^* = (\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\boldsymbol{Y} \ \text{and} \
\Sigma^* = \frac{1}{n-m-1}(\boldsymbol{Y}-\boldsymbol{X}\boldsymbol{B}^*)'(\boldsymbol{Y}-\boldsymbol{X}\boldsymbol{B}^*)
\]
For the "standard theory" prior the posterior of \(\Sigma\) is an inverse
Wishart distribution with \(df = n - m\) when \(k=2\) and \(df=n-m-1\) when \(k
= 1\) \citep{crawfordComparingSingleCase2011,
tiaoBayesianEstimationMultivariate1964} and scale matrix \((n-m-1)\Sigma^*\).
For each iteration we generate an estimate of \(\Sigma\) from this distribution.
For the "calibrated" prior (only used for \(k=2\)) the steps described in
\ref{BSDT} are followed and we use \(\pmb{S}^* = (n-m-2)\pmb{S}/(n-m-1)\) as the
scale matrix when we sample from an inverse Wishart distribution on \((n-m-2)\)
degrees of freedom, where \(\pmb{S} = (n-m-1)\Sigma^*\). The generated
observation from either method is denoted \(\hat{\Sigma}_{(i)}\).
We have

\begin{equation}
\hat{\Sigma}_{(i)}=[\hat{s}^2_{(i)}] \ \text{when} \ k=1 \ \text{and} \
\hat{\Sigma}_{(i)} =
\begin{bmatrix}
\hat{s}^2_{1(i)} & \hat{s}_{12(i)}  \\
\hat{s}_{12(i)}  & \hat{s}^2_{2(i)}
\end{bmatrix} \ \text{when} \ k=2 \
\label{eq:sigma}
\end{equation}

\(\boldsymbol{B}^*\) will be an \((m+1) \times k\) matrix. We turn this into a \(k(m + 1) \times 1\)
vector by concatenating the colums of regression coefficients in \(\boldsymbol{B}^*\), such
that \(\boldsymbol{B}^*_{\text{vec}}=(\boldsymbol{\beta}^{*'}_1, ...,\boldsymbol{\beta}^{*'}_k)'\).
Then this vector is the data estimate for
\(\boldsymbol{B}_{\text{vec}}=(\boldsymbol{\beta}'_1, ...,\boldsymbol{\beta}'_k)'\).
Take the kronecker product \(\boldsymbol{\hat\Sigma}_{(i)} \otimes (\boldsymbol{X'X})^{-1}\) and
denote this \(\boldsymbol{\Lambda}_{(i)}\). Then the posterior of \(\boldsymbol{B}_{\text{vec}}|\boldsymbol{\hat\Sigma}_{(i)}\)
is a multivariate normal distribution with mean vector \(\boldsymbol{B}^*_{\text{vec}}\)
and variance-covariance matrix \(\boldsymbol{\Lambda}_{(i)}\). Draw a random value
from this distribution such that
\(\hat{\boldsymbol{B}}^*_{\text{vec(i)}} =(\hat{\boldsymbol{\beta}}'_{1(i)}, ...,\hat{\boldsymbol{\beta}}'_{k(i)})'\)
and \(\hat{\boldsymbol{B}}_{(i)} = (\hat{\boldsymbol{\beta}}_{1(i)}, ...,\hat{\boldsymbol{\beta}}_{k(i)})\),
where each \(\hat{\boldsymbol{\beta}}_{j(i)}\) is a vector of length \(m+1\) and \(\hat{\boldsymbol{B}}_{(i)}\)
an \(m+1 \times k\) matrix.

Let \(\boldsymbol{x}^*\) be a vector of the case's values on the covariates.
The conditional values for the case on the tasks of interest
is then
\[
\hat{\boldsymbol{\mu}}_{(i)} = \hat{\boldsymbol{B}}_{(i)}\boldsymbol{x}^*
\]

We now estimate the effect size of the case using the conditional
estimations derived above. Depending on whether we are testing for a
deficit or a discrepancy (i.e., whether \(k=1\) or \(k=2\)) the calculations
of the effect sizes differ. \citet{crawfordComparingSingleCase2011} call these effect
sizes \(Z_{CCC}\) and \(Z_{DCCC}\) for a deficit and a discrepancy respectively.
They are similar to \(Z_{CC}\) (Equation \eqref{eq:zcc}) and \(Z_{DCC}\) (Equation \eqref{eq:PJ}), however,
the extra \(C\) in the subscript indicates that they are conditional on
covariates. Denote \(y^*_1\) and \(y^*_2\) as the case's scores on the two variates
of interest. Given that we want to estimate deficiency of a case on \(Y_1\), we
calculate
\begin{equation}
\hat{Z}_{CCC(i)} = \frac{y^*_1-\hat{\mu}_{(i)}}{\hat{s}^2_{(i)}}
\label{eq:zccc}
\end{equation}
For \(Z_{DCCC}\) we have two conditional means which will be denoted
\(\hat{\mu}_{1(i)}\) and \(\hat{\mu}_{2(i)}\) for \(Y_1\) and \(Y_2\)
respectively. We then calculate
\begin{equation}
\hat{Z}_{DCCC(i)} = \frac{\frac{y^*_1-\hat{\mu}_{1(i)}}{\hat{s}_{1(i)}}-\frac{y^*_2-\hat{\mu}_{2(i)}}
{\hat{s}_{2(i)}}}{\sqrt{2-2\hat{\rho}_{12(i)}}}
\label{eq:zdccc}
\end{equation}
Where \(\hat{s}_{1(i)}\) and \(\hat{s}_{2(i)}\) are conditional
standard deviations obtained from \(\hat{\Sigma}_{(i)}\) in Equation \eqref{eq:sigma}.
The conditional correlation between \(Y_1\) and \(Y_2\) in the denominator is given by
\[\hat{\rho}_{12(i)} = \frac{\hat{s}_{12(i)}}{\sqrt{\hat{s}^2_{1(i)}\hat{s}^2_{2(i)}}}\]

We then find the tail area under the standard normal distribution that is less or greater than
\(\hat{Z}_{CCC(i)}\) or \(\hat{Z}_{DCCC(i)}\) depending on the problem at hand and the alternative
hypothesis specified. Denote the value obtained \(\hat{p}_{(i)}\) which then
is an estimate of \(p\). If testing for a deficit we would have:
\[
\hat{p}_{(i)}= \Phi(\hat{Z}_{CCC(i)})
\]

Iterating these steps a large number of times will yield a distribution of
\(\hat{p}\), the mean of which is taken as the point estimate of \(p\). This
estimate is used for significance testing and if multiplied by 100 it will
give an estimate of the percentage of controls expected to exhibit a more
extreme score than the case. If repeated e.g., 1000 times, the
25th smallest and 25th largest \(\hat{p}_{(i)}\) is the lower and upper boundaries of
the 95\% credible interval for \(p\).

To obtain a point estimate of \(Z_{CCC}\) and \(Z_{DCCC}\) we use Equation \eqref{eq:zccc}
and \eqref{eq:zdccc}, but use the conditional means, standard deviations
and correlation calculated directly from the control sample.
The \(1-\alpha\) credible intervals for these
effect sizes are given by the \(\alpha/2\) and \(1-\alpha/2\) quantiles
of the sampled distribution of \(Z_{CCC}\) or \(Z_{DCCC}\). That is, just as
for \(p\), if repeated e.g., 1000 times, the
25th smallest and 25th largest value of \(\hat{Z}_{CCC(i)}\) or \(\hat{Z}_{DCCC(i)}\)
is the lower and upper boundaries of
the 95\% credible interval for \(Z_{CCC}\) or \(Z_{DCCC}\).


# The \pkg{singcar} package {short-title="The singcar package" #section3}

## Functions for testing abnormality

The main functions in the package are presented in Table \@ref(tab:funtab).


\begin{longtable}[]{@{\extracolsep{\fill}}llll@{}}
\tabularnewline
\toprule
\begin{minipage}[b]{(\columnwidth - 3\tabcolsep) * \real{0.32}}\raggedright
Name\strut
\end{minipage} & \begin{minipage}[b]{(\columnwidth - 3\tabcolsep) * \real{0.15}}\raggedright
Function\strut
\end{minipage} & \begin{minipage}[b]{(\columnwidth - 3\tabcolsep) * \real{0.29}}\raggedright
Source\strut
\end{minipage} & \begin{minipage}[b]{(\columnwidth - 3\tabcolsep) * \real{0.18}}\raggedright
Reference\strut
\end{minipage}\tabularnewline
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{(\columnwidth - 3\tabcolsep) * \real{0.32}}\raggedright
Name\strut
\end{minipage} & \begin{minipage}[b]{(\columnwidth - 3\tabcolsep) * \real{0.15}}\raggedright
Function\strut
\end{minipage} & \begin{minipage}[b]{(\columnwidth - 3\tabcolsep) * \real{0.29}}\raggedright
Source\strut
\end{minipage} & \begin{minipage}[b]{(\columnwidth - 3\tabcolsep) * \real{0.18}}\raggedright
Reference\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.32}}\raggedright
Test of deficit\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.15}}\raggedright
\code{TD()}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.29}}\raggedright
\citet{crawfordComparingIndividualTest1998}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.18}}\raggedright
Equation \eqref{eq:TD}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.32}}\raggedright
Bayesian test of deficit\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.15}}\raggedright
\code{BTD()}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.29}}\raggedright
\citet{crawfordComparisonSingleCase2007}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.18}}\raggedright
Section \ref{BTD}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.32}}\raggedright
Bayesian test of deficit with covariates\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.15}}\raggedright
\code{BTD_cov()}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.29}}\raggedright
\citet{crawfordComparingSingleCase2011}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.18}}\raggedright
Section \ref{cov}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.32}}\raggedright
Unstandardised difference test\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.15}}\raggedright
\code{UDT()}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.29}}\raggedright
\citet{crawfordTestingSuspectedImpairments2005}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.18}}\raggedright
Equation \eqref{eq:UDT}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.32}}\raggedright
Revised standardised difference test\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.15}}\raggedright
\code{RSDT()}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.29}}\raggedright
\citet{crawfordTestingSuspectedImpairments2005}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.18}}\raggedright
Equation \eqref{eq:RSDT}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.32}}\raggedright
Bayesian standardised difference test\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.15}}\raggedright
\code{BSDT()}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.29}}\raggedright
\citet{crawfordComparisonSingleCase2007}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.18}}\raggedright
Section \ref{BSDT}\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.32}}\raggedright
Bayesian standardised difference test with covariates\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.15}}\raggedright
\code{BSDT_cov()}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.29}}\raggedright
\citet{crawfordComparingSingleCase2011}\strut
\end{minipage} & \begin{minipage}[t]{(\columnwidth - 3\tabcolsep) * \real{0.18}}\raggedright
Section \ref{cov}\strut
\end{minipage}\tabularnewline
\bottomrule
\captionsetup{width=0.95\textwidth}
\caption{Main functions for significance testing of abnormality when using small control samples 
and principal relevant reference in prior literature.}
\label{tab:funtab}
\end{longtable}

To facilitate meta-analyses of studies using case-control comparisons,
all functions in Table \@ref(tab:funtab) can take both summary statistics as
input as well as raw data. The output will be a list set to class \code{"htest"},
for which the generic function \code{print} have a method. To use the package,
begin by installing and loading it:

```{r loadpack, eval=FALSE}
install.packages("singcar")
library("singcar")
```

## Example from neuropsychology

The package comes with the dataset \code{size_weight_illusion}, a
neuropsychological dataset from an investigation of the size-weight illusion in
DF, a patient with visual form agnosia following bilateral lesions to the
lateral occipital complex \citep{hassanSizeweightIllusionVisual2020}. The size-weight illusion
is a perceptual phenomenon in which smaller objects are perceived as heavier
during lifting than larger objects of equal weight \citep{buckinghamGettingGripHeaviness2014}.
The illusion implies that sensory cues about object size affect the perception
of weight. It has been suggested that patient DF does not experience this
illusion in the same way as the healthy population when only visual cues about
object size are available \citep{dijkermanVisuomotorPerformancePatient2004a}. In contrast, when kinaesthetic (tactile) cues are
provided it is suggested that DF's experience of the illusion is unaffected
by her brain damage. In other words, patient DF was expected to have a deficit
in the visual size-weight illusion and exhibit an abnormally large discrepancy
(dissociation) between visual and kinaesthetic size-weight illusions. The
dataset consists of data from patient DF and 28 control participants, with the
variables sex, age and visual as well as kinaesthetic size-weight illusion. The
measure of the size-weight illusion is a scaled measure expressing the number of
grams weight difference perceived per cubic cm of volume change. Below follows
examples of how to analyse this dataset using the tests provided in
\pkg{singcar}.

```{r datahead}
head(size_weight_illusion)
```


### Testing for a deficit

The simplest way to test for an abnormality on a single variate is
to use the frequentist test of deficit. Start by extracting patient (patient DF
is the first observation) and control data from the relevant variate, in this
case the visual size-weight illusion:
```{r VSWI}
PAT_VSWI <- size_weight_illusion[1, "V_SWI"]
CON_VSWI <- size_weight_illusion[-1, "V_SWI"] 
```
Using the function \code{TD()} we then apply the formula in Equation \eqref{eq:TD}.
The argument \code{conf_int_spec} specifies how fine grained the search
algorithm for the confidence interval should be. The arguments \code{sd}
and \code{sample_size} can be given if the test should be based on summary
statistics rather than raw data, the \code{controls} argument should then be the
mean of the control sample.
```{r TD}
TD(case = PAT_VSWI,
   controls = CON_VSWI,
   sd = NULL,
   sample_size = NULL,
   alternative = "less",
   conf_int = TRUE,
   conf_level = 0.95,
   conf_int_spec = 0.01,
   na.rm = FALSE)
```
This can similarly be tested with the Bayesian analogue which has a very similar syntax.
This test yields an output that converges to that of TD as the argument for the
number of iterations (\code{iter}) increase. The degrees of freedom shown in the
output below is the degrees of freedom for the \(\chi^2\) distribution from which
we sample \(\psi\), as described in Section \ref{BTD}.
```{r BTD}
set.seed(42)
BTD(case = PAT_VSWI,
    controls = CON_VSWI,
    sd = NULL,
    sample_size = NULL,
    alternative = "less",
    int_level = 0.95,
    iter = 10000,
    na.rm = FALSE)
```

If the control sample for a study is not appropriately matched to the case on
variables such as, for example, age or education level it is appropriate to use tests
that account for this by allowing for the inclusion of covariates. Including
theoretically sound covariates is often a good idea. \citet{crawfordComparingSingleCase2011}
recommends however to only include a covariate if it correlates \(\geq 0.3\) with
the variate of interest, because of the loss of degrees of freedom.

The function \code{BTD_cov()} allows for the inclusion of covariates and therefore to
assess the patient on the task of interest by essentially comparing him/her to
the controls with the same score on the covariate. Even though the correlation
between age and visual size-weight illusion is \(< 0.3\) it is included here as
a coviariate for demonstrative purposes. Start again by extracting the
scores on the covariate for the patient and for the control participants.
```{r covariate}
PAT_age <- size_weight_illusion[1, "YRS"] 
CON_age <- size_weight_illusion[-1, "YRS"]
```
Since \code{BTD_cov()} is somewhat computationally intense, the number of
iterations has been reduced in the example compared to \code{BTD()}. For actual analysis the number
of iterations should be based on required precision. It should be noted that
there is no restriction on the number of covariates used as long as \(n > m+1\). If more than one
covariate is used the case scores should be given as a vector of values
and the control scores should be given as a data frame or matrix with the
same number of columns as the number of values in the covariate vector of the case.

If summary statistics are used instead of raw data, the argument \code{use_sumstats}
must be set to \code{TRUE} and the correlation matrix for the covariates
and variate of interest must be given as well as the sample size. In addition,
the \code{control_covar} argument must be supplied as an \(m \times 2\)
matrix or data frame giving the mean of the covariate(s) in the first column
and the standard deviation in the second. The degrees of freedom shown in the
output below is the degrees of freedom for the inverse Wishart distribution from which
we sample \(\psi\), as described in Section \ref{cov}.
```{r BTD-cov}
BTD_cov(case_task = PAT_VSWI,
        case_covar = PAT_age,
        control_task = CON_VSWI,
        control_covar = CON_age,
        alternative = "less",
        int_level = 0.95,
        iter = 1000,
        use_sumstats = FALSE,
        cor_mat = NULL,
        sample_size = NULL)
```

### Testing for a dissociation

For assessing abnormal discrepancy between two variates the simplest
function to use is the unstandardised difference test, Equation \eqref{eq:UDT}.
This test is in \pkg{singcar} called by the function \code{UDT()}. However, one
should use this only if the variates are known to come from equivalent distributions.
Otherwise, tests that can evaluate standardised scores
without inflating Type I errors should be used. In the frequentist framework the
appropriate test for this is the RSDT, Equation \eqref{eq:RSDT}. In this example we wish
to estimate and test the abnormality of the discrepancy between visual
and kinaesthetic size-weight illusion in patient DF. That is, we want to compare the difference
between the variates exhibited by the patient and the distribution of differences
in the healthy control sample. Again, start by extracting
the patient and control scores for the second variate of interest.
```{r KSWI}
PAT_KSWI <- size_weight_illusion[1, "K_SWI"] 
CON_KSWI <- size_weight_illusion[-1, "K_SWI"] 
```
Using the function \code{RSDT()} we then apply the formula in
Equation \eqref{eq:RSDT}. This test is most often used two-sided due to the fact the the
sign of the discrepancy solely depends on the order of the input.
This function does, however, not provide any confidence intervals. The syntax of
\code{UDT()} is very similar to that of \code{RSDT()}, the main difference being
options for confidence intervals as shown for \code{TD()}. If summary statistics
are used then the additional argument \code{r_ab}, which is the sample
correlation, must be set as well as the sample size, standard deviation and mean
for both variates.
```{r RSDT}
RSDT(case_a = PAT_VSWI,
     case_b = PAT_KSWI,
     controls_a = CON_VSWI,
     controls_b = CON_KSWI,
     sd_a = NULL,
     sd_b = NULL,
     sample_size = NULL,
     r_ab = NULL,
     alternative = "two.sided",
     na.rm = FALSE)
```

The Bayesian analogue of this test is recommended over the RSDT
 because it keeps a better control
over Type I errors when the case exhibits extreme deficits on both variates but
no discrepancy between them \citep{crawfordComparisonSingleCase2007}.
The syntax of the two functions is similar, but the \code{BSDT()} comes with more
optional arguments. For example, one has the option of applying this test
without standardising the variates by setting the argument \code{unstandardised}
to \code{TRUE}. The output then converges to that of the frequentist UDT.
Furthermore, one can choose between priors. Setting the
argument \code{calibrated} to \code{FALSE} specifies the use of the "standard
theory" prior. If left to the default (\code{TRUE}) an accept-reject algorithm
is deployed for each simulation of \(\Sigma\), as described in Section
\ref{BSDT}. This default prior has been shown to have better frequentist
properties for estimating \(\rho\) and differences between means in a bivariate
normal distributions \citep{bergerObjectivePriorsBivariate2008} and is therefore
the default and recommended choice.
```{r BSDT}
BSDT(case_a = PAT_VSWI,
     case_b = PAT_KSWI,
     controls_a = CON_VSWI,
     controls_b = CON_KSWI,
     sd_a = NULL,
     sd_b = NULL,
     sample_size = NULL,
     r_ab = NULL,
     alternative = "two.sided",
     int_level = 0.95,
     iter = 10000,
     unstandardised = FALSE,
     calibrated = TRUE,
     na.rm = FALSE)
```

If analysing discrepancies between variates in the presence of covariates the syntax is slightly
different, requiring that one specifies the case's scores on the variates of
interest as a vector and the control scores as a data frame or matrix.
One has the option to choose between the "calibrated" and "standard theory"
prior, just as for \code{BSDT()}, where \code{calibrated = TRUE} is the
recommended and default behaviour. If using summary statistics
\code{use_sumstats} must be set to \code{TRUE} and the summary input should be
supplied to the arguments \code{control_tasks}/\code{control_covar} as data
frames or matrices with the means of each variable represented by the first
column and the standard deviations by the second. The case's scores should
be supplied as vectors.

```{r BSDT_cov}
BSDT_cov(case_tasks = c(PAT_VSWI, PAT_KSWI),
         case_covar = PAT_age,
         control_tasks = cbind(CON_VSWI, CON_KSWI),
         control_covar = CON_age,
         alternative = "two.sided",
         int_level = 0.95,
         calibrated = TRUE,
         iter = 1000,
         use_sumstats = FALSE,
         cor_mat = NULL,
         sample_size = NULL)
```

### Power calculators

A further capacity of \pkg{singcar} is that it can be used to calculate
statistical power of the tests. The notion of power when comparing cases to
control samples have been somewhat overlooked for this class of statistical tests.
In recent work \citep{mcintoshPowerCalculationsSinglecase2020}, we argued that,
even though power is inherently limited in this paradigm, a priori calculations are
still useful for study design and interpretation in neuropsychological and other applications.
Calculating power for the test of deficit is similar to calculating power for
any \(t\) test and can be done analytically.
\begin{equation} power = 1 - \beta =
T_{n-1}\left(t_{\alpha, \ n-1} \Bigg\rvert \frac{x^* - \overline{x}}{\sigma
\sqrt{\frac{n+1}{n}}}\right) \label{eq:TDpower} 
\end{equation}
Where \(T_{n-1}(.\rvert \theta)\) is the cumulative distribution function for the
non-central \(t\) distribution with \(n-1\) degrees of freedom and non-centrality
parameter \(\frac{y^* - \overline{y}}{\sigma \sqrt{\frac{n+1}{n}}}\) (i.e., TD, Equation
\eqref{eq:TD}) and \(t_{\alpha, \ n-1}\) is the \(\alpha\) quantile of the \emph{central}
\(t\) distribution on \(n-1\) degrees of freedom (note that this is for a one-sided
test). For the unstandardised difference test power is calculated in an
analogous way by putting Equation \eqref{eq:UDT} as the non-centrality parameter. Deriving
power for the other functions in an analytic manner is however not possible (the
RSDT is only approximately \(t\) distributed) and a Monte Carlo approach has been
used for these tests. To call any power calculator in the package one simply
uses the function names with \code{_power} added as a suffix.

So, for example, to calculate power for the test of deficit we call \code{TD_power()}.
The expected case score and either sample size or desired power must
be supplied. The mean and standard deviation of the control sample
can also be specified with the arguments \code{mean} and \code{sd}.
If not, they take the default values of 0 and 1 respectively so
that the case score is interpreted as distance from the mean
in standard deviations. A conventional \(\alpha\)-level of
\(0.05\) is assumed if nothing else is supplied. The alternative
hypothesis can also be specified by the argument \code{alternative}:
specify \code{"less"} (default) or \code{"greater"} for a one-tailed test, specify 
\code{"two.sided"} for a two-tailed test.
```{r TD-power}
TD_power(case = 70,
         mean = 100,
         sd = 15,
         sample_size = 16,
         power = NULL,
         alternative = "less",
         alpha = 0.05,
         spec = 0.005)
```

\code{TD_power()} can also be used to calculate required sample size for a
desired level of power. For example, if we specify a desired power level of 0.6,
leave \code{sample_size} to its default and let the rest of the arguments
be as in the previous example, we see from the output of the function that
power will not increase more than 0.5\% for any additional participant after a sample
size of 15. That is, the algorithm stops searching
when this level of specificity has been reached and we are nearing the
asymptotic maximum power for this effect size. We can increase the specificity
by lowering the \code{spec} argument.
```{r TDpowersize}
TD_power(case = 70,
         mean = 100,
         sd = 15,
         sample_size = NULL,
         power = 0.6,
         alternative = "less",
         alpha = 0.05,
         spec = 0.005)
```

Power calculators for the Bayesian tests of deficit cannot calculate
required sample size. This is because they rely on simulation methods
to estimate approximate power and deploying a search algorithm to find the required sample
size for a given level of power would be computationally too intense. The syntax
is otherwise relatively similar to that of \code{TD_power()}. For \code{BTD_power()}
we have the two extra arguments \code{nsim} and \code{iter}, indicating the number
of simulations used in the power function and by \code{BTD()}, respectively.
```{r BTDpower}
BTD_power(case = 70,
         mean = 100,
         sd = 15,
         sample_size = 15,
         alternative = "less",
         alpha = 0.05,
         nsim = 1000,
         iter = 1000)
```

The only difference in syntax of \code{BTD_cov_power()} is due to the inclusion of covariates.
The variate of interest must be specified as a vector where the first element
gives the mean and the second the standard deviation in the argument \code{control_task}.
The covariates can be specified similarly or as an \(m \times 2\) matrix where the first
column gives the means of each covariate and the second column gives the standard
deviations. The correlation matrix of the variates must be given as well. In the example
below, power is evaluated for a test taking two covariates into account, both with a mean
of 0 and a standard deviation of 1. The correlation is specified as a \(3\times 3\)
matrix with pairwise correlations of $0.3$. The default settings include only one
covariate having a $0.3$ correlation with the variate of interest.
This function is computationally intense and hence, the number
of simulations has, for the example below, been decreased.
```{r BTDcovpower}
covars <- matrix(c(0, 1,
                   0, 1), ncol = 2, byrow = TRUE)
BTD_cov_power(case = -2,
              case_cov = c(0.2, -0.6),
              control_task = c(0, 1),
              control_covar = covars,
              cor_mat = diag(3) + 0.3 - diag(c(0.3, 0.3, 0.3)),
              sample_size = 15,
              alternative = "less",
              alpha = 0.05,
              nsim = 100,
              iter = 100)
```

For the difference tests one must supply the expected case scores on both
variates as well as sample size. The means and standard deviations of the
control sample can also be specified. If unspecified, they take on the default values of
0 and 1 respectively, so that the expected case scores are interpreted as
distances from the means in standard deviations. \code{RSDT_power()},
\code{BSDT_power()} and \code{UDT_power()} additionally require an estimate of
the sample correlation between the variates of interest, \code{r_ab}. If this is
not specified a correlation of 0.5 is assumed by default. For
\code{BSDT_cov_power()} the correlation matrix between the variates of interest
and the covariates must instead be supplied (i.e., at least a \(3\times3\) matrix
where the first correlation is the correlation between the variates of
interest).

The alternative hypothesis is by default assumed to be
\code{"two.sided"} since the direction of the effect is dependent on the
order of the inputs, but can be specified to be \code{"less"} or
\code{"greater"} as well. The syntax is similar for all three functions but with small
differences. For \code{UDT_power()} one can request required sample size for a
desired power, as for \code{TD_power()}. Calculators for the Bayesian tests
have the extra argument \code{calibrated} as to be able to specify the prior.
\code{BSDT_cov_power()} requires input in the same format as \code{BTD_cov_power()}
for both \code{control_tasks} and \code{control_covar}. The two examples
below demonstrate usage for \code{RSDT_power()} and \code{BSDT_cov_power()}.
```{r RSDTpower}
RSDT_power(case_a = 70,
           case_b = 55,
           mean_a = 100,
           mean_b = 50,
           sd_a = 15,
           sd_b = 10,
           r_ab = 0.5,
           sample_size = 15,
           alternative = "two.sided",
           alpha = 0.05,
           nsim = 1000)
```

```{r BSDTcovpower}
cor_mat <- matrix(c(1,   0.5, 0.6,
                    0.5,   1, 0.3,
                    0.6, 0.3,   1), ncol = 3, byrow = TRUE)
BSDT_cov_power(case_tasks = c(70, 55),
               case_cov = 65,
               control_tasks = matrix(c(100, 15,
                                        50, 10), ncol = 2, byrow = TRUE),
               control_covar = c(50, 25),
               cor_mat = cor_mat,
               sample_size = 15,
               alternative = "two.sided",
               alpha = 0.05,
               nsim = 100,
               iter = 100,
               calibrated = TRUE)
```

# Summary

The \pkg{singcar} package for \proglang{R} has been outlined and
its main functionalities described. The package consists of methods for
estimating abnormality of a single case when compared to a population estimated
by a small sample. Methods for estimating abnormality on a single variate have been
described as well as methods for estimating abnormality of the difference
between two variates. Historically, the use of these tests has mainly been
within the field of neuropsychology, but their potential applicability is far
wider, extending to other areas of psychology and perhaps even to completely new
fields.

In neuropsychology the methods developed by John Crawford and Paul Garthwaite
are frequently used, especially the test of deficit, Equation \eqref{eq:TD}, and
the revised standardised difference test, Equation \eqref{eq:RSDT}. However, the
Bayesian standardised difference test, which outperforms the RSDT regarding
control of Type I errors, has not gained as much traction. Neither have the more
flexible Bayesian methods allowing for the inclusion of covariates.
It is hoped that by providing them in a documented package for a
popular language such as \proglang{R}, they will receive further uptake.

The methods described have been developed for keeping transparent control over
Type I errors, but power calculators have been implemented in the
package as well. Consideration of power can assist researchers in study design
and in setting realistic expectations for what these types of statistical
hypothesis tests can achieve \citep{mcintoshPowerCalculationsSinglecase2020}.
